Java Sorting Algorithms Comparison

### 2015-02-23 Programming Project B/paper.tex
added a graph and other shit too

%% BEGIN PROCEDURE
\section{Procedure}
An insertion sort can be implemented in a multitude of languages using the pseudocode provided in Algorithm 1.
\newline
\textbf{Insertion Sort Pre-Condition}: A is a non-empty array of data with a natural order.
\newline
\textbf{Insertion Sort Post-Condition}: A' is a permutation of A (containing all the same elements) in strictly non-decreasing order.
\begin{algorithm}
\caption {\textsc{Insertion-Sort}(A)}
\label{algo:insertionsort}
\begin{algorithmic}[1]
\Procedure{Insertion-Sort}{A}
\If{$A.length < 2$}
\State{\Return{$A$}}
\EndIf
\State{$i = 2$}
\While{$i$ upto $A.length$}
	\State{$key = A[i]$}
	\State{$j = i - 1$}
	\While{$j$ downto $1$ and $key < A[j]$}
		\State{$A[j + 1] = A[j]$}
		\State{$j = j - 1$}
	\EndWhile
	\State{$A[j+1] = key$}
	\State{$i = i + 1$}
\EndWhile
\State{\Return{$A$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Insertion Sort Outer-Loop Invariant}: The subarray A'[1 ... i - 1] contains all the same elements as the subarray A[1 .. i - 1].
\newline 
\textbf{Insertion Sort Outer-Loop Initialization}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same one element.
\newline
\textbf{Insertion Sort Outer-Loop Maintenance}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same elements, although they may be in different orders.
\newline
\textbf{Insertion Sort Outer-Loop Termination}: When the outer-loop terminates, i = A.length, which implies that the entire array has been traversed and the guard has been negated. The negation of the guard implies that A'[1 ... i - 1] contains all the elements in A[1 ... i - 1].
\newline
\newline
\textbf{Insertion Sort Inner-Loop Invariant}: A'[1 ... j] is sorted in strictly non-decreasing order.
\newline
\textbf{Insertion Sort Inner-Loop Initialization}: Before the first iteration of the loop, j = 1, meaning the subarray A'[1 ... j] contains exactly one element, which is already sorted.
\newline
\textbf{Insertion Sort Inner-Loop Maintenance}: At the beginning of each iteration of the loop the inner-loop invariant holds because j counts down from i, and A'[j+1] is swapped with A'[j] only if A'[j+1] is less than A[j].
\newline
\textbf{Insertion Sort Inner-Loop Termination}: The negation of the guard implies that j = A.length and that A'[1 ... j] has been entirely traversed and sorted in strictly non-decreasing order, which maintains the inner-loop invariant. 
\newline
\newline
\textbf{Insertion Sort Conclusion}: The termination of both the inner and outer loops implies that the entire array has been traversed, A' is a permutation of A containing all the same elements in strictly non-decreasing order. This satisfies the post condition.
\newline
\newline

@@

\textbf{Merge Sort First While Loop Invariant}: Left contains i elements, all of which can be found in A[0 ... i upto mid].

@@

\textbf{Merge Together First While Loop Invariant}: Combined contains x number of elements where x is the sum of i and y and those elements are contained in left[0 ... i] or right[0 ... y] in stricly non-decreasing order.

@@

\textbf{Merge Together Second While Loop Invariant}: Combined contains x number of elements where $x >= i$ and those elements are contained in left[0 ... i] in stricly non-decreasing order.

@@

\textbf{Merge Together Third While Loop Invariant}: Combined contains x number of elements where x is greather than or equal to y and those elements are contained in right[0 ... y] in stricly non-decreasing order.

@@

%% BEGIN EXPERIMENTAL ANALYSIS
\section{Experimental Analysis}
The insertion sort demonstrated in Algorithm 1, 2, and 3 was implemented in Java and executed on an HP SpectreXT TouchSmart with a 4 core Intel i7 processor clocked at 1.9GHz running Ubuntu Gnome 14.10 64-bit.
%% TODO: Include Graphs here

@@

%%TODO: Add shit about merge sort here
%%TODO: Add shit about heap sort here
%% END EXPERIMENTAL ANALYSIS

%% BEGIN CONCLUSION
\section{Conclusions}

%% END CONCLUSION

>>>>

%% BEGIN PROCEDURE
\section{Procedure}
An insertion sort can be implemented in a multitude of languages using the pseudocode provided in Algorithm 1.
\newline
\textbf{Insertion Sort Pre-Condition}: A is a non-empty array of data with a natural order.
\newline
\textbf{Insertion Sort Post-Condition}: A' is a permutation of A (containing all the same elements) in strictly non-decreasing order.
\begin{algorithm}
\caption {\textsc{Insertion-Sort}(A)}
\label{algo:insertionsort}
\begin{algorithmic}[1]
\Procedure{Insertion-Sort}{A}
\If{$A.length < 2$}
\State{\Return{$A$}}
\EndIf
\State{$i = 2$}
\While{$i$ upto $A.length$}
	\State{$key = A[i]$}
	\State{$j = i - 1$}
	\While{$j$ downto $1$ and $key < A[j]$}
		\State{$A[j + 1] = A[j]$}
		\State{$j = j - 1$}
	\EndWhile
	\State{$A[j+1] = key$}
	\State{$i = i + 1$}
\EndWhile
\State{\Return{$A$}}
\EndProcedure
\end{algorithmic}
\end{algorithm}
\newline
\textbf{Insertion Sort Outer-Loop Invariant}: The subarray A'[1 ... i - 1] contains all the same elements as the subarray A[1 ... i - 1].
\newline 
\textbf{Insertion Sort Outer-Loop Initialization}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same one element.
\newline
\textbf{Insertion Sort Outer-Loop Maintenance}: The outer-loop invariant holds because A'[1 ... i - 1] and A[1 ... i - 1] both contain the same elements, although they may be in different orders.
\newline
\textbf{Insertion Sort Outer-Loop Termination}: When the outer-loop terminates, i = A.length, which implies that the entire array has been traversed and the guard has been negated. The negation of the guard implies that A'[1 ... i - 1] contains all the elements in A[1 ... i - 1].
\newline
\newline
\textbf{Insertion Sort Inner-Loop Invariant}: A'[1 ... j] is sorted in strictly non-decreasing order.
\newline
\textbf{Insertion Sort Inner-Loop Initialization}: Before the first iteration of the loop, j = 1, meaning the subarray A'[1 ... j] contains exactly one element, which is already sorted.
\newline
\textbf{Insertion Sort Inner-Loop Maintenance}: At the beginning of each iteration of the loop the inner-loop invariant holds because j counts down from i, and A'[j+1] is swapped with A'[j] only if A'[j+1] is less than A[j].
\newline
\textbf{Insertion Sort Inner-Loop Termination}: The negation of the guard implies that j = A.length and that A'[1 ... j] has been entirely traversed and sorted in strictly non-decreasing order, which maintains the inner-loop invariant. 
\newline
\newline
\textbf{Insertion Sort Conclusion}: The termination of both the inner and outer loops implies that the entire array has been traversed, A' is a permutation of A containing all the same elements in strictly non-decreasing order. This satisfies the post condition.
\newline
\newline

@@

\textbf{Merge Sort First While Loop Invariant}: Left contains i elements, all of which can be found in A[1 ... i upto mid].

@@

\textbf{Merge Together First While Loop Invariant}: Combined contains x number of elements where x is the sum of i and y and those elements are contained in left[1 ... i] or right[1 ... y] in stricly non-decreasing order.

@@

\textbf{Merge Together Second While Loop Invariant}: Combined contains x number of elements where $x >= i$ and those elements are contained in left[1 ... i] in stricly non-decreasing order.

@@

\textbf{Merge Together Third While Loop Invariant}: Combined contains x number of elements where x is greather than or equal to y and those elements are contained in right[1 ... y] in stricly non-decreasing order.

@@

%% BEGIN EXPERIMENTAL ANALYSIS
\section{Experimental Analysis}
The insertion sort demonstrated in Algorithm 1, 2, and 3 was implemented in Java and executed on an HP SpectreXT TouchSmart with a 4 core Intel i7 processor clocked at 1.9GHz running Ubuntu Gnome 14.10 64-bit.
 \begin{figure}[!]
\begin{center}
\includegraphics[scale=.70]{test-results.pdf}
\end{center}
\captionof{figure}{Merge, Heap and Insertion Sort Run-Time Comparison}
\label{fig:comparisonfig}
\end{figure}

@@

The expected growth of merge sort as the number of elements (n) grows large can be represented as $ \theta(n*\lg(n)) $. At $n_0$ the sort took 0.0002 milliseconds to complete, however, at $n_1$ the algorithm took 0.0001 milliseconds to complete. The discrepency can be attributed to superfolous commands being executed by the system. As $n$ grows larger the time to complete correlates with the expected running time. It is expected for lower values of $n$ to not correlate as nicely with the expected running time because the system constant will have a larger impact on the smaller values of $n$. It is important to note that merge sort ran much faster than insertion sort at large values of n such as $n_{1*10^6} $ and was able to process $n_{1*10^7} $ where insertion sort crashed the system. The best case running time for this implementation of merge sort is $ \theta(1) $ and is elicited only when $ n <= 1 $ because if $n <= 1$ the array is vacuously sorted and can be returned immediately without executing any loop, and run at a constant time.

Heap sort is very similar to merge sort in that its expected running time is $ \theta(n*\lg(n)) $. At lower values of $n$ heap sort ran marginally quicker than merge sort, however, as $n$ grew larger merge sort began to run marginally faster than heap sort. This difference is neglagable because of superfolous commands being executed by the system and other running processes. For all intensive purposes merge sort and heap sort have the same run time and the major difference between the two is the system constant. For $n_0$ heap sort completed in 0.0002 milliseconds and for $n_1$ heap sort completed in 0.0003 milliseconds. In both merge sort and insertion sort the sort completes before any loop is executed if $n <= 1$ however in this implementation of heap sort the algorithm runs through the entire sort. We can tell from this that the constant time is 0.0002 milliseconds, which correlates nicely with both merge and insertion sort. $ \theta(n*\lg(n)) $ is the both the best and worst case run-time for heap sort.
%% END EXPERIMENTAL ANALYSIS

%% BEGIN CONCLUSION
\section{Conclusions}
With a running time of $ \theta(n*\lg(n)) $ both heap sort and merge sort perform much faster than insertion sort at a running time of $ \theta(n^2) $, especially when $n$ grows large. While faster comptuers will be able to execute insertion sort faster, as $n$ grows large, even on a slower machine, both merge sort and heap sort will perform faster. Because it is easier to setup and maintain, insertion sort is a good choice when dealing with small sets of data that don't go beyond $n_{1*10^5}$, however, if data is expected to grow larger than that, either heap or merge sort is the best choice.
%% END CONCLUSION
